How not to constructivize cohomology

### 2018-10-13 paper-flabby-objects.tex
gen pdf
Note slides almost finished

\title{Flabby and injective objects in toposes}
\author{Ingo Blechschmidt}
\address{Max Planck Institute for Mathematics in the Sciences \\
Inselstraße 22 \\
04103 Leipzig, Germany}
\email{ingo.blechschmidt@mis.mpg.de}

@@

@@

% XXX: In Einleitung interne Sprache erinnern.

@@

The notion of flabby objects originates from the notion of flabby shaves and is
therefore closely connected to Grothendieck toposes. It is therefore instructive
to study flabby objects in elementary toposes which are not Grothendieck
toposes, away from their original conceptual home. We begin this study with
establishing the following observations on flabby objects in the effective
topos. We follow the terminology of~\cite{hyland:effective-topos}.

>>>>

\title{Flabby and injective objects in toposes}
\author{Ingo Blechschmidt}
\address{Università di Verona \\
Department of Computer Science \\
Strada le Grazie 15 \\
37134 Verona, Italy}
\email{iblech@speicherleck.de}

@@

{\centering \textbf{Rough draft, comments welcome! iblech@speicherleck.de}\par}
\bigskip

@@

XXX: introduction
% XXX: In Einleitung interne Sprache erinnern.

@@

The notion of flabby objects originates from the notion of flabby shaves and is
therefore closely connected to Grothendieck toposes. Hence it is instructive
to study flabby objects in elementary toposes which are not Grothendieck
toposes, away from their original conceptual home. We begin this study with
establishing the following observations on flabby objects in the effective
topos. We follow the terminology of~\cite{hyland:effective-topos}.

### 2018-10-13 slides-pssl104.tex
Note slides almost finished

  \jnote{3-}{In constructive mathematics, the condition for a set to be flabby
  is nontrivial. We'll later see that the condition is also interesting for
  classical mathematics, if interpreted internally to suitable toposes.}

@@

  Given a space~$X$, we can hope that we can write~$X$ as the
  codomain of a continuous map~$f : X \to B$, in such a way that the base space~$B$
  and the fibers of~$f$ are in some sense easy to understand. In particular, we
  could ask whether the cohomology of~$X$ can be computed from the cohomology
  of~$B$ and the cohomology of the fibers.

  The answer, given by Jean Leray in the 1940s, is: Yes, we can, but the
  framework of singular cohomology is too rigid. For a positive answer we have
  to generalize to sheaf cohomology. And thus, the notion of sheaves was born.}
\end{frame}}

@@

  \jnote{1-}{An injective resolution is a sequence of sheaves and modules and
  linear morphisms as indicated such that the sequence is \emph{exact}
  (the kernel of any outgoing morphism equals the image of the respective incoming
  morphism) and such that the sheaves~$I^n$ are injective (a notion recalled
  below).

@@

  There are also \emph{reduction techniques} which propose interesting deals. For
  instance, there is a technique which allows us to pretend that reduced rings
  are Noetherian and in fact fields -- if in return we switch from classical
  reasoning to constructive. This particular technique has been used to turn
  the slightly convoluted multi-page proof of Grothendieck's generic
  freeness lemma into a simple one-paragraph argument.

@@

  A basic application is the following. Any student in algebraic geometry
  needs, at some point in her life, compute the cohomology of projective space.
  At a later point she needs to compute higher direct images
  along~$\mathbb{P}^n_S \to S$, where~$\mathbb{P}^n_S$ is a relative version of
  projective space. Since higher direct images are just internal
  sheaf cohomology, she can in fact skip the second computation.}
\end{frame}

@@

  \jnote{1-}{Since we cannot show the existence of enough injective sheaves of
  modules (or even just plain modules) constructively, the definition of sheaf
  cohomology using injective resolutions doesn't work in a constructive
  setting.}
\end{frame}

\begin{frame}{Flabbiness as an organizing principle}
%  \begin{block}{}
%    A set~$M$ is \hil{flabby} iff any \pointthisbelow{<1>}{subterminal}{$\forall x,y
%    \in K\_ x = y$}~$K \subseteq M$ is a \pointthisbelow{<1>}{subsingleton}{$\exists
%    a \in M\_ \forall x \in K\_ x = a$}.
%  \end{block}
%  \bigskip
%  \bigskip
%  \justifying

%  \textbf{Proposition.} Let~$E$ be a sheaf over a locale~$X$. Then~$E$ is flabby
%  iff it is a flabby set from the point of view of~$\Sh(X)$.
%
%  \textbf{Proof.} Routine/Zorn's lemma.

  \justifying\small
  \textbf{Proposition.} Let~$M$ be a sheaf of modules over a locale~$X$.
  Then~$M$ is injective iff it is injective from the point of view of~$\Sh(X)$.

@@

  Since~$M$ is injective, there exists a dotted map~$\bar g \in E$.
  We have~$K \subseteq \{ \bar g \}$. \qed
\end{frame}

@@

  \textbf{Question.} Are there enough flabby modules in the ef{}fective topos?
\end{frame}

@@

  \begin{itemize}
    \item Basic homological algebra is entirely constructive.
    \item There are algorithms for computing cohomology [Barakat, \ldots].
    \item Čech methods work constructively, even in a synthetic context.
  \end{itemize}
\end{frame}
\addtocounter{framenumber}{-1}

>>>>

  \jnote{3-}{In constructive mathematics, the condition for a set to be flabby
  is nontrivial. We'll see later that the condition is also interesting for
  classical mathematics, if interpreted internally to suitable toposes.}

@@

  Given a space~$X$, we can hope that we can write~$X$ as the
  codomain of a continuous map~$f : X \to B$, in such a way that the base space~$B$
  and the fibers of~$f$ are in some sense easy to understand. In such a situation we
  could ask whether the cohomology of~$X$ can be computed from the cohomology
  of~$B$ and the cohomology of the fibers.

  The answer, given by Jean Leray in the 1940s, is: Yes, we can, but the
  framework of singular cohomology is too rigid for this task. For a positive answer we have
  to generalize to sheaf cohomology. And thus, the notion of sheaves was born.}
\end{frame}}

@@

  \jnote{1-}{An injective resolution is a sequence of sheaves of modules and
  linear morphisms as indicated such that the sequence is \emph{exact}
  (the kernel of any outgoing morphism equals the image of the respective incoming
  morphism) and such that the sheaves~$I^n$ are injective (a notion recalled
  below).

@@

  There are also \emph{reduction techniques} which propose interesting deals. For
  instance, there is a technique which allows us to pretend that a reduced ring
  is Noetherian and in fact a field -- if in return we switch from classical
  reasoning to constructive. This particular technique has been used to turn
  the slightly convoluted multi-page proof of Grothendieck's generic
  freeness lemma into a simple one-paragraph argument.

@@

  A basic application is the following. Any student in algebraic geometry
  needs, at some point in her life, to compute the cohomology of projective space.
  At a later point she needs to compute higher direct images
  along~$\mathbb{P}^n_S \to S$, where~$\mathbb{P}^n_S$ is a relative version of
  projective space. Since higher direct images are just internal
  sheaf cohomology, she can in fact skip the second computation.}
\end{frame}

@@

  \jnote{1-}{Since we cannot show the existence of enough injective sheaves of
  modules (or even just plain modules) constructively, the definition of sheaf
  cohomology using injective resolutions doesn't work in a constructive
  setting. Classically it's known that \emph{flabby resolutions} can also be
  used to compute sheaf cohomology. There are more flabby sheaves than injective
  ones, they have better stability properties (flabby sheaves are preserved
  under pushforward) and the axiom of choice is not needed to construct flabby
  resolutions (the standard proof uses only the law of excluded middle, and not
  Zorn's lemma). Hence it seems reasonable to base a constructively sensible
  definition of sheaf cohomology on flabby resolutions. We tried to do so, and
  failed.

  Assuming Zorn's lemma, the notion of a flabby sheaf is a local notion,
  meaning that a sheaf is flabby if and only if its restrictions to every
  member of an open covering are, but this fact is not obvious from the
  definition. In contrast, the notion that a sheaf~$E$ is flabby from the
  internal point of view is local without any assumptions (as is any
  internal notion), hence maybe we should consider adopting internal flabbiness as the
  official definition of flabbiness. Its external translation is:

  A sheaf~$E$ is flabby from the internal point of view if and only if for any
  local section~$s \in E(U)$, there is an open covering~$X = \bigcup_i U_i$
  such that for all~$i$, the section~$s$ extends to a section on~$U \cup U_i$.}
\end{frame}

\begin{frame}{Flabbiness as an organizing principle}
  \justifying\small
  \textbf{Proposition.} Let~$M$ be a sheaf of modules over a locale~$X$.
  Then~$M$ is injective iff it is injective from the point of view of~$\Sh(X)$.

@@

  Since~$M$ is injective, there exists a dotted map~$\bar g \in E$.
  We have~$K \subseteq \{ \bar g \}$. \qed

  \jnote{1-}{The notion of being flabby from the internal point of view turns
  out to have valuable organizing power. For instance, both of the
  following statements can be proven by first verifying that a certain sheaf is
  internally flabby (which can be done entirely constructively) and then
  appealing to Zorn's lemma in order to obtain a global section of that sheaf.

  \begin{itemize}\justifying
    \item Flabby sheaves are acyclic for the global sections functor: Let~$0
    \to E \to F \to G \to 0$ be a short exact sequence of sheaves of modules
    and let~$E$ be flabby. Then the sequence remains exact after taking global
    sections.

    (Verify that the sheaf of local preimages of a given section~$s \in G(X)$ is
    flabby.)

    \item Internally injective modules are externally injective.

    (See proof on the slide.)
  \end{itemize}}
\end{frame}

@@

  \textbf{Question.} Are there enough flabby modules in the ef{}fective topos?

  \jnote{1-}{The notion of flabby sets was conceived to model the notion of
  flabby sheaves and is therefore closely connected to Grothendieck toposes.
  Hence it is instructive to study flabby objects in elementary toposes which
  are not Grothendieck toposes, away from their original conceptual home.

  In particular, we hope to prove the conjecture that the statement ``any
  module embeds into a flabby module'' is not constructively provable by
  verifying that it doesn't hold in the effective topos.

  To this end, the slide displays two results.

  Details on the self-referential construction~``$K \defeq \{ f(a_K) \}$'' are in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-flabby-objects.pdf}{this
  draft paper}.

  A module is~$\neg\neg$-separated if and only if~$\neg\neg(x = 0)$ implies~$x = 0$.}
\end{frame}

@@

  \begin{itemize}
    \item Basic homological algebra is entirely constructive.
    \item There are algorithms for computing cohomology [Barakat, \ldots].
    \item Čech methods work constructively, even in a synthetic context.
  \end{itemize}

  \jnote{1-}{Even if we could constructively prove that there are enough flabby
  modules, there is still the problem that the proof that flabby
  sheaves of modules are acyclic for the global sections functor (appears to)
  require Zorn's lemma.}
\end{frame}
\addtocounter{framenumber}{-1}
